---
title: "Cox models"
author: "Zhaoqi Liu"
date: "2/26/2021"
output: pdf_document
---

```{r}
bcdf<-readRDS("bcdf_final.rds")
bcdf$Race<-ifelse(bcdf$Race=="Black",0,1) #convert to numeric variable
bcdf<-as.data.frame(bcdf)
```

```{r}
library(ggplot2)
library(survival)
library(glmnet)
library(forestplot)
```


### For Overall Survival:

#### Using 2 predictors: average neculus count and average cytomembrane count of LSR
```{r}
attach(bcdf)
cor(AveNCount,AveCMCount)
plot(AveNCount,AveCMCount)
detach(bcdf)
ggplot(data=bcdf,aes(x=Survival_month,y=AveTCount))+geom_point(aes(color=as.factor(Survival)))
ggplot(data=bcdf,aes(x=Survival_month,y=AveNCount))+geom_point(aes(color=as.factor(Survival)))
```


Simple Cox Model and check the proportional-hazards assusmption 
```{r}
#model.add<-coxph(Surv(Survival_month,Survival)~AveNCount+AveCMCount,data=bcdf)#colinearity
model.n<-coxph(Surv(Survival_month,Survival)~AveNCount,data=bcdf)
model.n
test.ph.n<-cox.zph(model.n)
ggcoxzph(test.ph.n)
model.cm<-coxph(Surv(Survival_month,Survival)~AveCMCount,data=bcdf)
model.cm
test.ph.cm<-cox.zph(model.cm)
ggcoxzph(test.ph.cm) #assumption roughly hold 
```

In model.add, no features are statistically significant based on the wald test.

```{r}
#logLik(model.add)
logLik(model.n)
logLik(model.cm)
#partial likelihood ratio test 
2*(-512.6753--513.2001)#2*l(full)-l(reduced) with df =2-1= 1
pchisq(1.0496,1,lower.tail = FALSE) #effect of AveNCount is not statistically significant when AveCMCount is included in the model
2*(-512.6753--512.9614)
pchisq(0.5722,1,lower.tail = FALSE) #effect of AveCMCount is not statistically significant when AveNCount is included in the model 

#null model
logLik(coxph(Surv(Survival_month,Survival)~1,data=bcdf))
pchisq(2*(-513.2001--515.4559),1,lower.tail = FALSE) #AveNCount byitself is significant , so is AveCMCount
AIC(coxph(Surv(Survival_month,Survival)~1,data=bcdf)) #null model AIC
```

```{r}
AIC(model.add)
AIC(model.cm) #best among the three 
AIC(model.n)
```


### check if there are unusual pattern 

```{r}
lapply(2:9,FUN=function(x){
  ggplot(data=bcdf,aes(x=Survival_month,y=AveTCount,color=unlist(bcdf[,x])))+
    geom_point()+
    labs(color=names(bcdf)[x])
})
```

#### Using 1 predictors: average total count of LSR 

```{r}
summary(coxph(Surv(Survival_month,Survival)~AveTCount,data=bcdf,ties="efron"))


model.t<-coxph(Surv(Survival_month,Survival)~AveTCount,data=bcdf,ties="efron")
model.t
test.ph.t<-cox.zph(model.t,transform="km")
ggcoxzph(test.ph.t)
# "discrete exact method"
```

Martingale residuals 
```{r}
result.coxph.0<-coxph(Surv(Survival_month, Survival)~1, data=bcdf)
rr.0<-residuals(result.coxph.0,type="martingale")

#assess the potential relationship of survival to AveTCount
smoothSEcurve <- function(yy, xx) {
# use after a call to "plot"
# fit a lowess curve and 95% confidence interval curve
# make list of x values
  xx.list <- min(xx) + ((0:100)/100)*(max(xx) - min(xx))
# Then fit loess function through the points (xx, yy)
# at the listed values
  yy.xx <- predict(loess(yy ~ xx), se=T,
  newdata=data.frame(xx=xx.list))
  lines(yy.xx$fit ~ xx.list, lwd=2)
  lines(yy.xx$fit -
  qt(0.975, yy.xx$df)*yy.xx$se.fit ~ xx.list, lty=2)
  lines(yy.xx$fit +
  qt(0.975, yy.xx$df)*yy.xx$se.fit ~ xx.list, lty=2)
}
plot(rr.0~bcdf$AveTCount)
smoothSEcurve(rr.0,bcdf$AveTCount) #fit a LOESS(locally weighted scatterplot smoothing) curve




```




apply the univariate coxph function to multiple covariate at once. 
```{r}
covariates <- c("AveNCount","AveCMCount","AveTCount")
univ_formulas <- sapply(covariates,
                        function(x) as.formula(paste('Surv(Survival_month, Survival)~', x)))
                        
univ_models <- lapply(univ_formulas, function(x){coxph(x, data = bcdf)})
# Extract data 
univ_results <- lapply(univ_models,
                       function(x){ 
                          x <- summary(x)
                          p.value<-signif(x$wald["pvalue"], digits=2)
                          wald.test<-signif(x$wald["test"], digits=2)
                          beta<-signif(x$coef[1], digits=2);#coeficient beta
                          HR <-signif(x$coef[2], digits=2);#exp(beta)
                          HR.confint.lower <- signif(x$conf.int[,"lower .95"], 2)
                          HR.confint.upper <- signif(x$conf.int[,"upper .95"],2)
                          #HR <- paste0(HR, " (", 
                          #             HR.confint.lower, "-", HR.confint.upper, ")")
                          res<-c(beta, HR, HR.confint.lower, 
                                 HR.confint.upper,wald.test, p.value)
                          names(res)<-c("beta", "HR",
                                        "HR.confint.lower","HR.confint.upper", 
                                        "wald.test", "p.value")
                          return(res)
                          #return(exp(cbind(coef(x),confint(x))))
                         })
res <- t(as.data.frame(univ_results, check.names = FALSE))
res
#as.data.frame(res)
```

Client request investigation: 
```{r}
summary(coxph(Surv(Survival_month,Survival)~AveTCount+ER+Race,data=bcdf,ties="efron"))
AIC(coxph(Surv(Survival_month,Survival)~AveTCount+ER+Race,data=bcdf,ties="efron"))

summary(coxph(Surv(Survival_month,Survival)~AveNCount+AveCMCount+ER+Race,data=bcdf,ties="efron"))
AIC(coxph(Surv(Survival_month,Survival)~AveTCount+AveCMCount+ER+Race,data=bcdf,ties="efron"))

ggsurvplot(surv_fit(Surv(Survival_month,Survival)~ER,data=kmpdf,group.by="Race"),data=kmpdf,pval=TRUE)
```

### Stepwise variable selection
```{r}
#forcing AveTCount in the model
model.All.coxph<-coxph(Surv(Survival_month, Survival)~Race+Age+ER+PR+HER2+Grade+Stage+AveTCount+AveNCount+AveCMCount,data=bcdf)
result.step<-step(model.All.coxph, 
                  scope = list(upper = ~Race+Age+ER+PR+HER2+Grade+Stage+AveTCount+AveCMCount+AveNCount, 
                               lower = ~AveTCount))  
result.step 
coxph(Surv(Survival_month, Survival)~Age+Grade+Stage+AveTCount,data=bcdf)
```

```{r}
# not force AveTCount in the data 
result.step<-step(model.All.coxph)
result.step 
```

```{r}
ggforest(result.step)
```


forest plot for the Cox model with multiple variables 
```{r}
#result.step 
coef.est <- c(0.044207, NA, NA,0, 0.433748, NA, NA, 0, 0.605995, NA,NA,0,1.200739, NA,0.841978)
se.est <- c(0.007857, NA,NA, 0, 0.283465, NA, NA, 0, 0.230007, NA, NA,0,0.225934,NA, 0.570057)
lower <- coef.est - 1.96*se.est
upper <- coef.est + 1.96*se.est
label.factors <- matrix(c(" Age","","", "HER2 negative", "HER2 positive", "", "",   "Grade low", "Grade high", "", "", "Stage  low", "Stage high", "", "LSR in nucleus"), ncol=1)
```

```{r}
forestplot(label.factors, coef.est, lower=lower, upper=upper,
boxsize=0.4, xticks=c(-0.5,0,0.5,1,1.5, 2),
txt_gp=fpTxtGp(label=gpar(cex=0.8), 
               ticks = gpar(cex=0.8),
               xlab=gpar(cex=0.8)), 
xlab="log hazard ratio",
col = fpColors(box = "royalblue",
                          line = "darkblue",
                          summary = "royalblue"))
```

```{r}

```


```{r}
cox.zph(result.step)
ggcoxzph(cox.zph(result.step))

#martingale residual 
rr.final<-residuals(result.step, type="martingale")
plot(rr.final~bcdf$AveTCount)
smoothSEcurve(rr.final,bcdf$AveTCount)
```



```{r}
set.seed(2021)
cv.lasso.fit<-cv.glmnet(x,y,family="cox",type.measure="C") # Harrell C index:  a higher C index means better prediction performance
plot(cv.lasso.fit)
cv.lasso.fit$lambda.min
```

```{r}
coef(lasso.fit,s=cv.lasso.fit$lambda.min)
coef(lasso.fit,s=0.05) #lambda = 0.05 only four features are selected 
```


```{r}
cox.lassovar<-coxph(Surv(Mod_Relapse_month,Mod_Relapse)~AveTCount+Race+Age+ER+HER2+Grade+Stage,data=bcdf)
```


### Forest Plot 

###For Relapse: 
Simple cox 
```{r}
coxph(Surv(Mod_Relapse_month,Mod_Relapse)~AveNCount,data=bcdf)
coxph(Surv(Mod_Relapse_month,Mod_Relapse)~AveCMCount,data=bcdf)
coxph(Surv(Mod_Relapse_month,Mod_Relapse)~AveTCount,data=bcdf)

cox.zph(coxph(Surv(Mod_Relapse_month,Mod_Relapse)~AveNCount,data=bcdf))
cox.zph(coxph(Surv(Mod_Relapse_month,Mod_Relapse)~AveCMCount,data=bcdf))
cox.zph(coxph(Surv(Mod_Relapse_month,Mod_Relapse)~AveTCount,data=bcdf))
ggcoxzph(cox.zph(coxph(Surv(Mod_Relapse_month,Mod_Relapse)~AveTCount,data=bcdf)))
```

```{r}
covariates <- c("AveNCount","AveCMCount","AveTCount")
univ_formulas <- sapply(covariates,
                        function(x) as.formula(paste('Surv(Mod_Relapse_month, Mod_Relapse)~', x)))
                        
univ_models <- lapply(univ_formulas, function(x){coxph(x, data = bcdf)})
# Extract data 
univ_results <- lapply(univ_models,
                       function(x){ 
                          x <- summary(x)
                          p.value<-signif(x$wald["pvalue"], digits=2)
                          wald.test<-signif(x$wald["test"], digits=2)
                          beta<-signif(x$coef[1], digits=2);#coeficient beta
                          HR <-signif(x$coef[2], digits=2);#exp(beta)
                          HR.confint.lower <- signif(x$conf.int[,"lower .95"], 2)
                          HR.confint.upper <- signif(x$conf.int[,"upper .95"],2)
                          #HR <- paste0(HR, " (", 
                          #             HR.confint.lower, "-", HR.confint.upper, ")")
                          res<-c(beta, HR, HR.confint.lower, 
                                 HR.confint.upper,wald.test, p.value)
                          names(res)<-c("beta", "HR",
                                        "HR.confint.lower","HR.confint.upper", 
                                        "wald.test", "p.value")
                          return(res)
                          #return(exp(cbind(coef(x),confint(x))))
                         })
res <- t(as.data.frame(univ_results, check.names = FALSE))
res
```


